---
title: "Quant Problem Set 1 (1)"
format: html
editor: visual
---


## Problem 1 
### (a)

```{r}
# Set a seed so the random sample can be reproduced
set.seed(1)

# Create a function that calculates the three estimates
variance_estimators <- function(x) {
  
  # Number of observations in the sample
  n <- length(x)
  
  # Variance estimator dividing by n
  var_n <- sum((x - mean(x))^2) / n
  
  # Sample variance dividing by n-1
  sample_var <- var(x)
  
  # Sample standard deviation
  sample_sd <- sd(x)
  
  # Return the three estimates
  c(var_n = var_n,
    sample_var = sample_var,
    sample_sd = sample_sd)
}

# Generate a sample of 10 observations from N(0, 2)
x <- rnorm(10, mean = 0, sd = 2)

# Check the sample
x

# Apply the function to the sample
variance_estimators(x)
```



### (b)

```{r}
# Set a seed so the simulation can be reproduced
set.seed(1)

# Simulate 1,000 samples of 5 observations
simulation_results <- replicate(1000, {
  x <- rnorm(5, mean = 0, sd = 2)
  
  # Apply the function from part (a)
  variance_estimators(x)
})

# Convert the simulation results into a data frame
# Each row represents one simulated sample
results <- data.frame(
  sample = 1:1000,
  t(simulation_results)
)

# Check the first few rows
head(results)

# Calculate the mean of the two variance estimators
mean_var_n <- mean(results$var_n)
mean_sample_var <- mean(results$sample_var)

# Display the means
mean_var_n
mean_sample_var

# True population variance
true_var <- 4

# Create common breaks so the two histograms can be compared
max_value <- max(results$var_n, results$sample_var)
common_breaks <- seq(0, max_value, length.out = 31)

# Plot the sampling distribution of the divide-by-n estimator
hist(results$var_n,
     breaks = common_breaks,
     freq = FALSE,
     col = "lightgray",
     main = "Sampling Distributions of Variance Estimators",
     xlab = "Estimated Variance")

# Add the sampling distribution of the sample variance
hist(results$sample_var,
     breaks = common_breaks,
     freq = FALSE,
     add = TRUE,
     col = NA,
     border = "red",
     lwd = 2)

# Mark the true variance
abline(v = true_var,
       lwd = 2)

# Mark the mean of the divide-by-n estimator
abline(v = mean_var_n,
       col = "blue",
       lwd = 2,
       lty = 2)

# Mark the mean of the sample variance
abline(v = mean_sample_var,
       col = "red",
       lwd = 2,
       lty = 2)

# Add a legend
legend("topright",
       legend = c("Divide by n", "Sample variance",
                  "True variance", "Mean: divide by n",
                  "Mean: sample variance"),
       col = c("gray", "red", "black", "blue", "red"),
       lty = c(1, 1, 1, 2, 2),
       lwd = 2,
       cex = 0.8)
```

Answer: The divide-by-n estimator is biased downward. Its mean across the 1,000 samples is approximately 3.358, which is below the true variance of 4. The sample variance (dividing by n - 1) has a mean of about 4.198, which is relatively close to 4, so it is approximately unbiased. The divide-by-n estimator tends to underestimate the true variance because the deviations are measured from the sample mean rather than the true population mean.



### (c)

```{r}
# Set a seed so the simulation can be reproduced
set.seed(1)

# Sample sizes to compare
sample_sizes <- c(3, 5, 10, 30, 100)

# True population variance
true_var <- 4

# Create an empty data frame to store the bias results
bias_results <- data.frame()

# Repeat the simulation for each sample size
for (n in sample_sizes) {
  
  # Simulate 1,000 samples of size n
  simulation_results <- replicate(1000, {
    
    # Generate a sample from N(0, 2)
    x <- rnorm(n, mean = 0, sd = 2)
    
    # Apply the function from part (a)
    variance_estimators(x)
  })
  
  # Convert the results so that each row represents one sample
  simulation_df <- data.frame(t(simulation_results))
  
  # Calculate the estimated bias of the two variance estimators
  bias_var_n <- mean(simulation_df$var_n) - true_var
  bias_sample_var <- mean(simulation_df$sample_var) - true_var
  
  # Store the results
  bias_results <- rbind(
    bias_results,
    data.frame(
      sample_size = n,
      bias_var_n = bias_var_n,
      bias_sample_var = bias_sample_var
    )
  )
}

# Display the bias results
bias_results

# Plot the bias of the divide-by-n estimator
plot(bias_results$sample_size,
     bias_results$bias_var_n,
     type = "b",
     ylim = range(bias_results$bias_var_n,
                  bias_results$bias_sample_var,
                  0),
     xlab = "Sample Size",
     ylab = "Estimated Bias",
     main = "Bias of Variance Estimators")

# Add the bias of the sample variance
lines(bias_results$sample_size,
      bias_results$bias_sample_var,
      type = "b",
      col = "red")

# Add a horizontal line at zero bias
abline(h = 0,
       lty = 2)

# Add a legend
legend("bottomright",
       legend = c("Divide by n", "Sample variance"),
       col = c("black", "red"),
       lty = 1,
       pch = 1)
```



### (d)

```{r}
# Set a seed so the simulation can be reproduced
set.seed(1)

# Sample sizes to compare
sample_sizes <- c(3, 5, 10, 30, 100)

# True population standard deviation
true_sd <- 2

# Create an empty data frame to store the bias results
sd_bias_results <- data.frame()

# Repeat the simulation for each sample size
for (n in sample_sizes) {
  
  # Simulate 1,000 samples of size n
  simulation_results <- replicate(1000, {
    
    # Generate a sample from N(0, 2)
    x <- rnorm(n, mean = 0, sd = 2)
    
    # Apply the function from part (a)
    variance_estimators(x)
  })
  
  # Convert the results so that each row represents one sample
  simulation_df <- data.frame(t(simulation_results))
  
  # Calculate the estimated bias of the sample standard deviation
  bias_sample_sd <- mean(simulation_df$sample_sd) - true_sd
  
  # Store the result
  sd_bias_results <- rbind(
    sd_bias_results,
    data.frame(
      sample_size = n,
      bias_sample_sd = bias_sample_sd
    )
  )
}

# Display the estimated bias at each sample size
sd_bias_results
```

Answer: No, s is not an unbiased estimator of σ = 2. Its estimated bias is approximately -0.112 at n = 5, -0.067 at n = 10, -0.008 at n = 30, 0.001 at n = 100. The results show that s tends to underestimate σ, especially at small sample sizes and the bias gets smaller as n grows. This happens even though s² is unbiased for σ². Because taking the square root is a nonlinear transformation, unbiasedness does not necessarily survive a nonlinear transformation.



### (e)

```{r}
# Calculate the standard deviation of the 1,000 divide-by-n estimates (n = 5)
sd_var_n <- sd(results$var_n)

# Calculate the standard deviation of the 1,000 sample variance estimates (n = 5)
sd_sample_var <- sd(results$sample_var)

# Display the standard deviations
sd_var_n
sd_sample_var
```

Answer: The standard deviation of the 1,000 divide-by-n variance estimates is 2.316, while the standard deviation of the 1,000 sample variance estimates is 2.895. Therefore, the divide-by-n estimator is less variable. This is because it equals the sample variance multiplied by (n − 1)/n = 0.8, which shrinks the spread of the estimates.

However, this does not make it the better estimator. Estimator quality depends on both bias and variability. The divide-by-n estimator is downward biased, so its estimates are tightly clustered around the wrong value. In contrast, the sample variance is unbiased for the true population variance, even though it is more variable.



## Problem 2
### (a)

```{r}
# Create a function that generates one dataset from the true process
make_data <- function(n, rho) {
  
  # Generate z (political interest) from N(0, 1)
  z <- rnorm(n, mean = 0, sd = 1)
  
  # Generate u (the part of x unrelated to z) from N(0, 1)
  u <- rnorm(n, mean = 0, sd = 1)
  
  # Generate the error term from N(0, 1)
  epsilon <- rnorm(n, mean = 0, sd = 1)
  
  # Generate x (news consumption) so that its correlation with z is rho
  x <- rho * z + sqrt(1 - rho^2) * u
  
  # True parameter values
  beta0 <- 1
  beta1 <- 2
  beta2 <- 3
  
  # Generate y (political knowledge) from the true model
  y <- beta0 + beta1 * x + beta2 * z + epsilon
  
  # Return the data as a data frame
  data.frame(y, x, z)
}

# Set a seed so the results can be reproduced
set.seed(1)

# Generate 100 observations with rho = 0.5
d <- make_data(n = 100, rho = 0.5)

# Check the first few rows
head(d)

# Check that x and z are roughly N(0, 1)
mean(d$x)
sd(d$x)
mean(d$z)
sd(d$z)

# Check that the correlation between x and z is about 0.5
cor(d$x, d$z)
```

Answer: The sample means of x and z are close to 0 (0.02 and 0.11) and their standard deviations are close to 1 (0.94 and 0.90). The correlation between x and z is 0.475, close to the target of 0.5. Small gaps are expected with only 100 observations, so the data match the intended distributions.



### (b)

```{r}
# Fit the theorized model (omits z)
model_short <- lm(y ~ x, data = d)

# Fit the true model (includes z)
model_true <- lm(y ~ x + z, data = d)

# Estimated coefficients
coef(model_short)
coef(model_true)

# Coefficient tables with standard errors
summary(model_short)
summary(model_true)

# Confidence intervals
confint(model_short)
confint(model_true)

# Plot y against x
plot(d$x, d$y,
     xlab = "x (news consumption)",
     ylab = "y (political knowledge)",
     main = "y against x, n = 100, rho = 0.5")

# Add the fitted line from the theorized model
abline(model_short, col = "red", lwd = 2)

# Add the fitted line from the true model (holding z at 0)
abline(a = coef(model_true)["(Intercept)"],
       b = coef(model_true)["x"],
       col = "blue", lwd = 2)

# Add the true line (beta0 = 1, beta1 = 2)
abline(a = 1, b = 2, lty = 2, lwd = 2)

# Add a legend
legend("topleft",
       legend = c("y ~ x", "y ~ x + z", "True line"),
       col = c("red", "blue", "black"),
       lty = c(1, 1, 2),
       lwd = 2,
       cex = 0.8)
```

Answer: The model that leaves out z overestimates the effect of x. Its x coefficient is 3.32 (95% CI: 2.76 to 3.88), and the interval misses the true value of 2. The true model gets it right: 1.94 (95% CI: 1.69 to 2.19), which includes 2. The plot shows the same thing: the red line (y ~ x) is too steep, while the blue line (y ~ x + z) sits right on top of the dashed true line. This is omitted variable bias. People who read more news also tend to be more politically interested, and political interest raises knowledge. Without z, the model gives x credit for z's effect.



### (c)

```{r}
# Set a seed so the results can be reproduced
set.seed(1)

# Generate 1,000 observations with rho = 0.5
d_large <- make_data(n = 1000, rho = 0.5)

# Check the variables
mean(d_large$x)
sd(d_large$x)
mean(d_large$z)
sd(d_large$z)
cor(d_large$x, d_large$z)

# Fit both models
model_short_large <- lm(y ~ x, data = d_large)
model_true_large <- lm(y ~ x + z, data = d_large)

# Estimated coefficients, standard errors, and confidence intervals
coef(model_short_large)
coef(model_true_large)
summary(model_short_large)
summary(model_true_large)
confint(model_short_large)
confint(model_true_large)

# Plot y against x
plot(d_large$x, d_large$y,
     xlab = "x (news consumption)",
     ylab = "y (political knowledge)",
     main = "y against x, n = 1000, rho = 0.5")

# Add the fitted line from the theorized model
abline(model_short_large, col = "red", lwd = 2)

# Add the fitted line from the true model (holding z at 0)
abline(a = coef(model_true_large)["(Intercept)"],
       b = coef(model_true_large)["x"],
       col = "blue", lwd = 2)

# Add the true line
abline(a = 1, b = 2, lty = 2, lwd = 2)

# Add a legend
legend("topleft",
       legend = c("y ~ x", "y ~ x + z", "True line"),
       col = c("red", "blue", "black"),
       lty = c(1, 1, 2),
       lwd = 2,
       cex = 0.8)
```

Answer: More data did not resolve the bias. The x coefficient from the model without z is 3.54 (95% CI: 3.37 to 3.71), still far above the true value of 2, and the interval is now even further from 2 because the standard error shrank from 0.28 to 0.09. The true model (y ~ x + z) gives 2.03 (95% CI: 1.95 to 2.10), which includes 2.

A larger sample makes an estimate more precise, but it doesn't fix a misspecified model. Leaving out z keeps pushing the slope toward the wrong value (about 2 + 3 × 0.5 = 3.5), so the estimate just becomes a more confident wrong answer.


### (d)

```{r}
# Set a seed so the results can be reproduced
set.seed(1)

# Generate 1,000 observations with rho = 0
d_zero <- make_data(n = 1000, rho = 0)

# Check the variables
mean(d_zero$x)
sd(d_zero$x)
mean(d_zero$z)
sd(d_zero$z)
cor(d_zero$x, d_zero$z)

# Fit both models
model_short_zero <- lm(y ~ x, data = d_zero)
model_true_zero <- lm(y ~ x + z, data = d_zero)

# Estimated coefficients, standard errors, and confidence intervals
coef(model_short_zero)
coef(model_true_zero)
summary(model_short_zero)
summary(model_true_zero)
confint(model_short_zero)
confint(model_true_zero)

# Plot y against x
plot(d_zero$x, d_zero$y,
     xlab = "x (news consumption)",
     ylab = "y (political knowledge)",
     main = "y against x, n = 1000, rho = 0")

# Add the fitted line from the theorized model
abline(model_short_zero, col = "red", lwd = 2)

# Add the fitted line from the true model (holding z at 0)
abline(a = coef(model_true_zero)["(Intercept)"],
       b = coef(model_true_zero)["x"],
       col = "blue", lwd = 2)

# Add the true line
abline(a = 1, b = 2, lty = 2, lwd = 2)

# Add a legend
legend("topleft",
       legend = c("y ~ x", "y ~ x + z", "True line"),
       col = c("red", "blue", "black"),
       lty = c(1, 1, 2),
       lwd = 2,
       cex = 0.8)
```

Answer: With rho = 0, leaving out z no longer biases the estimate. The x coefficient from y ~ x is 2.04 (95% CI: 1.84 to 2.24), which includes the true value of 2 and matches the 2.02 from the true model. Omitted variable bias only appears when the left-out variable is related to both x and y. Here z affects y but is unrelated to x, so x doesn't pick up any of its effect.

This means we should include variables that affect y and are related to x, based on theory, since leaving them out biases the estimate and more data won't fix it. Since we never know rho in real data, we can't count on omitted variables being harmless.
